\begin{frame}
\frametitle{Course outline - Day 1}
Building Android
\begin{itemize}
\item Introduction to Android
\item Getting Android sources
\item Building and booting Android
\item Introduction to the Linux kernel
\item Compiling and booting the Linux kernel
\end{itemize}
Labs: download Android sources, compile them and boot them with the
Android emulator. Recompile the Linux kernel.
\end{frame}

\begin{frame}
\frametitle{Course outline - Day 2}
Android kernel, boot and filesystem details
\begin{itemize}
\item Android changes to the Linux kernel
\item Android bootloaders
\item Booting Android
\item Using ADB
\item Android filesystem
\end{itemize}
Labs: customize, compile and boot Android for the BeagleBone Black board.
\end{frame}

\begin{frame}
\frametitle{Course outline - Day 3}
Supporting a new product and customizing it
\begin{itemize}
\item Android build system. Add a new module and product.
\item Android native layer - Bionic, Toolbox, init, various daemons,
Dalvik, hardware abstraction, JNI...
\end{itemize}
Labs: Use ADB, create a new product, customize the product for the
BeagleBone Black board.
\end{frame}

\begin{frame}
\frametitle{Course outline - Day 4}
Android framework and applications
\begin{itemize}
\item Android framework for applications
\item Introduction to application development
\item Android packages
\item Advise and resources
\end{itemize}
Labs: compile an external library and a native application to control a
USB missile launcher. Create a JNI library and develop an Android
application to control the device.
\end{frame}
