\section{Creating and using an SDK}

\subsection{Introduction to the SDK}

\begin{frame}
  \frametitle{Overview}
  \begin{itemize}
    \item An SDK (Software Development Kit) is a set of tools allowing
      the development of applications for a given target (operating
      system, platform, environment\dots).
    \item It generally provides a set of of tools including:
      \begin{itemize}
        \item Compilers or cross-compilers.
        \item Linkers.
        \item Library headers.
        \item Debuggers.
        \item Custom utilities.
      \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Advantages}
  \begin{itemize}
    \item The Yocto Project is often used to build images for embedded
      targets.
      \begin{itemize}
        \item This often requires a special toolchain, to cross
          compile the software.
        \item Some libraries headers may be specific to the target
          and not available on the developers' computers.
      \end{itemize}
    \item A self-sufficient environment makes development easier and
      avoids many errors.
    \item Long manuals are not necessary, the only thing required is
      the SDK!
    \item Using the SDK to develop an application limits the risks of
      dependency issues when running it on the target.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{The Poky SDK}
  \begin{itemize}
    \item The Poky reference system is used to generate images, by
      building many applications and doing a lot configuration work.
      \begin{itemize}
        \item When developing an application, we only care about the
          application itself.
        \item We want to be able to develop, test and debug easily.
      \end{itemize}
    \item The Poky SDK is an application development SDK, which can be
      generated to provide a full environment compatible with the
      target.
    \item It includes a toolchain, libraries headers and all the
      needed tools.
    \item This SDK can be installed on any computer and is
      self-contained. The presence of Poky is not required for the SDK
      to fully work.
  \end{itemize}
\end{frame}

\subsection{Generating an SDK}

\begin{frame}
  \frametitle{Available SDKs}
  \begin{itemize}
    \item Two different SDKs can be generated:
      \begin{itemize}
        \item A generic SDK, including:
          \begin{itemize}
            \item A toolchain.
            \item Common tools.
            \item A collection of basic libraries.
          \end{itemize}
        \item An image-based SDK, including:
          \begin{itemize}
            \item The generic SDK.
            \item The sysroot matching the target root filesystem.
            \item Its toolchain is self-contained (linked to an
              SDK embedded libc).
          \end{itemize}
      \end{itemize}
    \item The SDKs generated with Poky are distributed in the form of a
      shell script.
    \item Executing this script extracts the tools and sets up the
      environment.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{The generic SDK}
  \begin{itemize}
    \item Mainly used for low-level development, where only the
      toolchain is needed:
      \begin{itemize}
        \item Bootloader development.
        \item Kernel development.
      \end{itemize}
    \item The recipe \code{meta-toolchain} generates this SDK:
      \begin{itemize}
        \item \code{bitbake meta-toolchain}
      \end{itemize}
    \item The generated script, containing all the tools for this SDK,
      is in:
      \begin{itemize}
        \item \code{build/tmp/deploy/sdk}
        \item Example:
          \code{poky-eglibc-x86_64-meta-toolchain-i586-toolchain-1.5.3.sh}
      \end{itemize}
    \item The SDK will be configured to be compatible with the
      specified \code{MACHINE}.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{The image-based SDK}
  \begin{itemize}
    \item Used to develop applications running on the target.
    \item One task is dedicated to the process. The task behavior
      can vary between the images.
      \begin{itemize}
        \item \code{populate_sdk}
      \end{itemize}
    \item To generate an SDK for \code{core-image-minimal}:
      \begin{itemize}
        \item \code{bitbake -c populate_sdk core-image-minimal}
      \end{itemize}
    \item The generated script, containing all the tools for this SDK,
      is in:
      \begin{itemize}
        \item \code{build/tmp/deploy/sdk}
        \item Example:
          \code{poky-eglibc-x86_64-core-image-minimal-i586-toolchain-1.5.3.sh}
      \end{itemize}
    \item The SDK will be configured to be compatible with the
      specified \code{MACHINE}.
  \end{itemize}
\end{frame}

\subsection{Use the SDK}

\begin{frame}[fragile]
  \frametitle{SDK format}
  \begin{itemize}
    \item Both SDKs are distributed as bash scripts.
    \item These scripts self extract themselves to install the
      toolchains and the files they provide.
    \item To install an SDK, retrieve the generated script and execute
      it.
      \begin{itemize}
        \item The script asks where to install the SDK. Defaults to
          \code{/opt/poky/<version>}
        \item Example: \code{/opt/poky/1.5.3}
        \begin{block}{}
          \begin{minted}[fontsize=\scriptsize]{console}
$ ./poky-eglibc-x86_64-meta-toolchain-i586-toolchain-1.5.3.sh
Enter target directory for SDK (default: /opt/poky/1.5.3):
You are about to install the SDK to "/opt/poky/1.5.3". Proceed[Y/n]?
Extracting SDK...done
Setting it up...done
SDK has been successfully set up and is ready to be used.
          \end{minted}
        \end{block}
      \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Use the SDK}
  \begin{itemize}
    \item To use the SDK, a script is available to set up the
      environment:
      \begin{block}{}
        \begin{minted}{console}
$ cd /opt/poky/1.5.3
$ source ./environment-setup-i586-poky-linux
        \end{minted}
      \end{block}
    \item The \code{PATH} is updated to take into account the binaries
      installed alongside the SDK.
    \item Environment variables are exported to help using the tools.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{SDK installation}
  \begin{description}
    \item[environment-setup-i586-poky-linux] Exports environment
      variables.
    \item[site-config-i586-poky-linux] Variables used during the
      toolchain creation
    \item[sysroots] SDK binaries, headers and libraries. Contains
      one directory for the host and one for the target.
    \item[version-i586-poky-linux] Version information.
  \end{description}
\end{frame}

\begin{frame}
  \frametitle{SDK environment variables}
  \begin{description}
    \item[CC] Full path to the C compiler binary.
    \item[CFLAGS] C flags, used by the C compiler.
    \item[CXX] C++ compiler.
    \item[CXXFLAGS] C++ flags, used by \code{CPP}
    \item[LD] Linker.
    \item[LDFLAGS] Link flags, used by the linker.
    \item[ARCH] For kernel compilation.
    \item[CROSS\_COMPILE] For kernel compilation.
    \item[GDB] SDK GNU Debugger.
    \item[OBJDUMP] SDK objdump.
  \end{description}
  \begin{itemize}
    \item To see the full list, open the environment script.
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Examples}
  \begin{itemize}
    \item To build an application for the target:
      \begin{block}{}
        \begin{minted}{console}
$ $CC -o example example.c
        \end{minted}
      \end{block}
    \item The \code{LDFLAGS} variables is set to be used with the C
      compiler (\code{gcc}).
      \begin{itemize}
        \item When building the Linux kernel, unset this variable.
          \begin{block}{}
            \begin{minted}{console}
$ unset LDFLAGS
$ make menuconfig
$ make
            \end{minted}
          \end{block}
      \end{itemize}
  \end{itemize}
\end{frame}

\subsection{Eclipse integration}

\begin{frame}
  \frametitle{The Yocto Project ADT}
  \begin{itemize}
    \item Eclipse can be used as a graphical interface to the Yocto
      SDK.
      \begin{itemize}
        \item To modify and develop recipes and applications.
        \item To compile packages and build images.
        \item To debug the system.
      \end{itemize}
    \item This is done thanks to an Eclipse Yocto plugin.
    \item The SDK and the Eclipse plugin is called the Yocto
      Application Development Toolkit.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Installation}
  \begin{itemize}
    \item The Eclipse Yocto plugin can be installed from a pre-built
      version available at:
      \begin{itemize}
        \item \code{http://downloads.yoctoproject.org/releases/eclipse-plugin/1.6.1/kepler}
      \end{itemize}
    \item Or it can be compiled from the official repository:
      \begin{itemize}
        \item \code{git://git.yoctoproject.org/eclipse-poky-kepler}
      \end{itemize}
    \item The full explanations to install and use the Eclipse Yocto
      plugin are given in the labs.
  \end{itemize}
\end{frame}
