\section{What's new in Buildroot?}

\begin{frame}{What's new in Buildroot}
  \begin{itemize}
  \item The major improvements in each release are summarized in the
    file named \code{CHANGES} in the Buildroot source tree
  \item Always mentions changes that could cause backward
    compatibility problems
  \item The following slides summarize the major new features added in
    each release between 2013.08 and 2015.08.
  \item All new Buildroot versions come with new packages, and many
    updates to the existing packages
    \begin{itemize}
    \item Such package additions and updates are not listed in the
      following slides.
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{In 2013.08 (1)}
  \begin{itemize}
  \item {\bf Architectures:} improved support for floating point on
    ARM and Thumb/Thumb2, support for ARM OABI removed
  \item {\bf Toolchains:}
    \begin{itemize}
    \item support added for Sourcery CodeBench ARM and MIPS 2013.05
    \item Linaro ARM and Aarch64 toolchains updated
    \item support added for the Arago ARMv5 and ARMv7 toolchains
    \item gcc 4.8.x version bumped
    \item support for installing both FDPIC and FLAT libraries on
      Blackfin
    \item support for uClibc 0.9.31 removed
    \item convert the internal toolchain backend to use the package
      infrastructure
    \item support added for eglibc in the internal toolchain backend
    \item toolchain components for the ARC architecture updated and
      gdb for ARC added.
    \item support for Blackfin in the internal toolchain fixed
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{In 2013.08 (2)}
  \begin{itemize}
  \item {\bf Defconfigs:} BeagleBone defconfig updated, new defconfig
    for CubieBoard, for Olimex mx233 Olinuxino, for Calao Systems
    TNY-A9G20-LPW
  \item {\bf Packages:}
    \begin{itemize}
    \item A number of packages have been fixed to use the
      \code{<pkg>_CONFIG_SCRIPTS} mechanism to get their
      \code{<pkg>-config} shell script installed and modified
      properly.
    \item Licensing information has been added to a number of
      packages.
    \item Use XZ tarballs for a number of packages
    \end{itemize}
  \item {\bf Noticeable package changes/additions}
    \begin{itemize}
    \item The glib2/libgtk2/webkit stack has been updated to recent versions.
    \item Support for Gstreamer 1.x has been added.
    \item OpenGL support for TI OMAP platforms has been added.
    \item OpenGL support for Allwinner platforms has been added.
    \item OpenMAX support for RaspberryPi has been added.
    \end{itemize}
  \item Top level menu names reordered and renamed for clarity
  \end{itemize}
\end{frame}

\begin{frame}{In 2013.11}
  \begin{itemize}
  \item {\bf Architectures:} Nios-II support, MIPS arch handling fixes
  \item {\bf Toolchains:}
    \begin{itemize}
    \item glibc support
    \item upstream uClibc fixes
    \item uClibc 0.9.31 for avr32
    \item internal crosstool-ng backend removed
    \item external musl toolchain support
    \item gcc 4.8.2
    \item updated Linaro external toolchains
    \item Fortran and objective-C support deprecated
    \item mudflap support
    \end{itemize}
  \item {\bf Bootloaders:} U-Boot: u-boot.imx support, version bumps
  \item {\bf Linux:} use kmod instead of module-init-tools
  \item {\bf System:} default to devtmpfs for /dev
  \item {\bf Infrastructure:} Infrastructure: Make 3.82 fixes, locales
    generation fixes, CVS download support, post-rsync hooks
  \end{itemize}
\end{frame}

\begin{frame}{In 2014.02}
  \begin{itemize}
  \item Support for external packages/defconfigs: \code{BR2_EXTERNAL}
  \item Cleanup of environment variable names for consistency
  \item {\bf Toolchain}: new Linaro and Sourcery Codebench
    toolchains. x86: Support for AMD Jaguar cores, SSE4.x, SH:
    SH2/SH3/SH3EB variants removed, Microblaze: Internal toolchain
    support
  \item {\bf Legal infrastructure:} Info is now split between host and
    target packages, large number of license annotations.
  \item {\bf Lua}: selection between lua 5.1 / 5.2, luarocks support
  \item {\bf Python:} package infrastructure, many new packages
  \item {\bf Defconfigs:} Armadeus APF51 and Zedboard added, apf27,
    apf28, beaglebone, microblaze, pandaboard, qemu, raspberry pi
    updated.
  \end{itemize}
\end{frame}

\begin{frame}{In 2014.05 (1)}
  \begin{itemize}
  \item {\bf Architectures:}
    \begin{itemize}
    \item Support for MIPS o32 ABI on MIPS-64 targets has
      been removed (too exotic)
    \item Support for the ARM A12 variant and Intel corei7
    \end{itemize}
  \item {\bf Defconfigs:} Minnowboard and Altera SoCkit added, QEMU
    updates.
  \item {\bf Bootloaders:} Grub2 and gummiboot support, syslinux
    support extended.
  \item Kconfig handling for minimum kernel headers version required
    for packages. Now packages needing specific kernel header features
    can specify these requirements in Kconfig.
  \end{itemize}
\end{frame}

\begin{frame}{In 2014.05 (2)}
  \begin{itemize}
  \item {\bf Toolchains:}
    \begin{itemize}
    \item GCC 4.9. Glibc 2.19.
    \item Support for the musl C
      library for internal and external toolchains.
    \item GCC 4.8-R3 support for ARC
    \item Internal toolchain support for Aarch64 and
      Microblaze
    \item Toolchain tuple vendor name can now be
      customized.
    \item Updated external Linaro ARM/Aarch64
      toolchains.
    \item Added external Linaro ARMEB toolchain.
    \item A GDB gdbinit file is now generated for external toolchains
      to automatically set the correct sysroot.
    \end{itemize}
  \item {\bf Infrastructure:}
    \begin{itemize}
    \item Support for (but disabled as it leads to
      unreproducible builds) toplevel parallel builds.
    \item Python package infrastructure extended to support Python 3.x
    \item Perl and virtual package infrastructure support added.
    \item \code{PRE_*_HOOKS} support for all build steps.
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{In 2014.08 (1)}
  \begin{itemize}
  \item {\bf Architectures:}
    \begin{itemize}
    \item Powerpc64 BE/LE added, AVR32 deprecated.
    \item Improved altivec / SPE /atomic instructions
      handling. Additional PowerPC CPU variants added.
    \end{itemize}
  \item {\bf Defconfigs:} Atmel SAMA5D3, Congatec QMX6, Lego ev3,
    TS-5x00, qemu-system-xtensa, qemu-aarch64-virt added. A number
    of tweaks to existing ones. lpc32xx defconfigs removed.
  \item {\bf Toolchain:}
    \begin{itemize}
    \item Microblaze support for internal musl toolchain.
    \item Default to GCC 4.8 for internal toolchain, remove deprecated
      4.3 and 4.6 versions.
    \item External CodeSourcery / Linaro toolchain updates
    \item Option to copy gconv libraries for external toolchains.
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{In 2014.08 (2)}
  \begin{itemize}
  \item {\bf Infrastructure:}
    \begin{itemize}
    \item \code{graph-depends} improvements
    \item Download handling is now done using helper scripts.
    \item Integrity of downloads can now be verified using hashes
    \item Legal-info: License info of local or overridden packages
      are saved as well. Toolchain packages are also taken into
      account.
    \item autotools: Static linking with libtool / v1.5 improvements
    \item Gettextize support, similar to autoreconf
    \item kconfig package infrastructure added
    \end{itemize}
  \item User manual restructured / reworked
  \end{itemize}
\end{frame}

\begin{frame}{In 2014.11}
  \begin{itemize}
  \item {\bf Toolchains:}
    \begin{itemize}
    \item Use \code{-mcpu} / \code{-march} instead of \code{-mtune}
    \item Support additional ARC and sparc variants
    \item Updated Code sourcery and Linaro external toolchains
    \end{itemize}
  \item {\bf Defconfigs:} Freescale iMX6DL SabreSD, Minnowboard MAX,
    QEMU powerpc64 pseries added and a number of updates to the
    existing configurations.
  \item {\bf Infrastructure:}
    \begin{itemize}
    \item Buildroot is now less noisy when built with the silent
      option (\code{make -s})
    \item A number of package infrastructure variables have been
      renamed from \code{*_OPT} to \code{*_OPTS} for consistency
    \item Option to choose what shell /bin/sh points to
    \end{itemize}
  \item {\bf Documentation:}
    \begin{itemize}
    \item Various updates to the user manual
    \item The asciidoc documentation handling has now been extended so
      it can be used by \code{BR2_EXTERNAL}
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{In 2015.02 (1)}
  \begin{itemize}
  \item {\bf Static/shared library handling reworked}
    \begin{itemize}
    \item This is now a tristate (shared only / shared and static /
      static only)
    \item Default is now shared only to speed up the
      build. \code{BR2_PREFER_STATIC_LIB} is now called
      \code{BR2_STATIC_LIBS}
    \end{itemize}
  \item {\bf Toolchain:}
    \begin{itemize}
    \item The toolchain (internal and external) will now warn when an
      unsafe library or header path is used
    \item If \code{BR2_COMPILER_PARANOID_UNSAFE_PATH} is enabled under
      build options this instead becomes an error.
    \end{itemize}
  \item {\bf Architectures:} Freescale E5500 and E6500 PowerPC support
    added, deprecated MIPS 1/2/3/4 support removed.
  \item {\bf Defconfigs:} Freescale p2020ds, MIPS creator CI20,
    Raspberrypi with DT, UDOO Quad
  \end{itemize}
\end{frame}

\begin{frame}{In 2015.02 (2)}
  \begin{itemize}
  \item \code{make <foo>_defconfig} now saves the path to the
    defconfig in the \code{.config}, so a \code{make savedefconfig}
    automatically updates it
  \item Infrastructure for packages using the Erlang rebar tool has
    been added.
  \item Hashes for a large number of packages have been added. Hashes
    are now checked for both target and host packages.
  \item The system menu now has an option to automatically configure a
    network interface through DHCP at bootup.
  \item The default filesystem skeleton now uses a separate
    \code{tmpfs} for \code{/run} instead of a symlink to \code{/tmp/}
    for security reasons / to protect against conflicts with user
    generated temporary files.
  \item \code{BR2_EXTERNAL} is now exported to post-build and
    post-image scripts.
  \end{itemize}
\end{frame}

\begin{frame}{In 2015.05 (1)}
  \begin{itemize}
  \item {\bf Architectures:}
    \begin{itemize}
    \item Removed AVR32 support, SuperH64 deprecated
    \item Added support for steamroller, corei7-avx and core-avx2 x86
      variants.
    \end{itemize}
  \item {\bf Toolchains:}
    \begin{itemize}
    \item IPv6 and Largefile support now enforced for
      uClibc. Corresponding Kconfig symbols removed.
    \item External CodeSourcery AMD64 2014.05 added
    \item musl-cross 1.1.6 added
    \item CodeSourcery SuperH 2 and Xilinx Microblaze v2/14.3 removed
    \item Distro-class external toolchains are now detected and
      blacklisted
    \item Internal toolchain support for Nios2 added, Blackfin
      removed.
    \item Aarch64 and sh musl support.
    \item uClibc-ng support added
    \item Libatomic is now handled for internal and external
      toolchains.
    \item Link time optimization (LTO) support.
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{In 2015.05 (2)}
  \begin{itemize}
  \item {\bf Defconfigs:} Freescale i.MX28 EVK, i.MX31 PDK and SABRE
    Auto, Raspberry Pi 2, RIoTboard
  \item {\bf Infrastructure:}
    \begin{itemize}
    \item Hashes for a large number of packages have
      been added.
    \item Missing hashes now stop the build unless
      explicitly disabled.
    \item Spaces and colons (:) are now supported in package versions.
    \item Dependencies can now be listed for the patch step
      (\code{<PKG>_PATCH_DEPENDENCIES}).
    \item Kconfig and Linux kernel extensions infrastructure has
      been added.
    \item Makedevs now has a recursive (r) option
    \item \code{external-deps}, \code{legal-info}, \code{source},
      \code{source-check} have been reimplemented using the package
      infrastructure, so their output/behaviour may differ from
      earlier (some packages were not included in the past).
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{In 2015.08}
  \begin{itemize}
  \item {\bf Architectures:} Minimal support for ARM Cortex-M3 and
    AArch64 big-endian.
  \item {\bf Toolchains:} Use {\em uClibc-ng} by default, add gcc 5.x
    support, update toolchain components
  \item {\bf Defconfigs:} VIA VAB-820/AMOS-820, OLimex OLinuxino A20
    Lime, many Atmel evaluation boards, ACME Systems Aria G25,
    WarPboard, Altera Cyclone 5 Development Board, Xilinx zc706, ARC
    AXS101 and AXS103
  \item {\bf Infrastructure:}
    \begin{itemize}
    \item Predictable permissions in the generated rootfs
    \item Support for {\em kconfig fragments}
    \item New \code{kernel-module} infrastructure
    \item Rework of the skeleton and init scripts packaging
    \item New \code{linux-tools} infrastructure in the \code{linux}
      package
    \item GCC version dependency mechanism
    \end{itemize}
  \item {\bf Filesystems:} Complete rework of the ISO9660 support.
  \end{itemize}
\end{frame}
