\subchapter{Bootloader optimizations}{Reduce bootloader execution time}

In this lab, we will run the final stage of boot time reduction:
\begin{itemize}
\item Improving the efficiency of the bootloader by optimizing its
usage
\item Recompiling the bootloader with the minimum set of options,
and even completely skip the second stage of the bootloader.
\end{itemize}

\section{Optimizing U-Boot usage}

By following the indications given in the lectures, start by optimizing
the way U-Boot is used.

At last, you can start by eliminating the infamous 2-second boot delay, something
you've surely been longing to do.

\section{Recompiling the bootloader}

It's now time to eliminate useless features in U-Boot. Go to
\code{~/boot-time-labs/bootloader/u-boot/} and run \code{make
menuconfig} to unselect features that we don't need in our system.

For the moment, don't touch the \code{SPL / TPL} options, as we will try
to use U-Boot's Falcon mode at the end.

In the same way you did when you reduced the kernel configuration,
do the changes {\bf progressively}, and even make backup copies of your
intermediate configurations (\code{.config} file). You will be glad you
did when you break U-Boot.

Once you have reached the minimum set of features, please measure boot
time and fill the below table:

\begin{tabular}{| l | l | r |}
  \hline
  Step & Duration & Description \\
  \hline
  \hline
  U-Boot SPL & & Between \code{U-Boot SPL 2021.10} and \code{U-Boot 2021.10} \\
  \hline
  U-Boot & & Between \code{U-Boot 2021.10} and \code{Starting kernel} \\
  \hline
  Kernel + Init scripts & & Between \code{Starting kernel} and \code{Starting ffmpeg} \\
  \hline
  Application & & Between \code{Starting ffmpeg} and \code{First frame decoded} \\
  \hline
  \hline
  Total & & \\
  \hline
\end{tabular}

\section{Using faster storage}

A last minute surprise: your instructor will give you new SD cards with
faster read performance, at least as fast as the Beagle Bone Black seems
to be able to go.

Why on earth didn't we use such SD cards right from the start of our
labs?

It's because slower storage acts as a magnifying glass (or as a slow
motion device) making it easier to observe elapsed time and the benefits
of our optimizations. If the storage was lightning fast, it would be
harder to appreciate speedups due to a small initramfs, for example.

So, edit the partition table of your new SD card, and create the
first partition in the same way as when you prepared your original SD
card. Then, copy the files over.

You can now go ahead and make tests again, and fill the table with your
latest results:

\begin{tabular}{| l | l | r |}
  \hline
  Step & Duration & Description \\
  \hline
  \hline
  U-Boot SPL & & Between \code{U-Boot SPL 2021.10} and \code{U-Boot 2021.10} \\
  \hline
  U-Boot & & Between \code{U-Boot 2021.10} and \code{Starting kernel} \\
  \hline
  Kernel + Init scripts & & Between \code{Starting kernel} and \code{Starting ffmpeg} \\
  \hline
  Application & & Between \code{Starting ffmpeg} and \code{First frame decoded} \\
  \hline
  \hline
  Total & & \\
  \hline
\end{tabular}

\section{Using U-Boot's {\em Falcon} mode}

It's now time to try U-Boot's capability to directly load the
Linux kernel from its first stage (SPL), instead of loading U-Boot.

What follows is based on U-Boot's own documentation in its sources:
\begin{itemize}
\item \projfile{u-boot}{doc/README.falcon} (generic details)
\item \projfile{u-boot}{board/ti/am335x/README} (specific details for boards with
      the am335x SoC)
\end{itemize}

The first thing to do is to generate a \code{uImage} file for the kernel
binary. This image file contains information that U-Boot uses to know a
few things about the kernel binary, most importantly the final load
address, but also the type of file (binary, script, environment file),
the target architecture and whether the binary is compressed or not.

This is called a {\em legacy image} for U-Boot. As you already know,
U-Boot can now boot a \code{zImage} file, but according to the Falcon
mode documentation, it does need a \code{uImage} file for SPL loading.

So, let's generate this file:
\begin{verbatim}
cd ~/boot-time-labs/kernel/linux/
make uImage LOADADDR=80008000
\end{verbatim}

Copy this \code{uImage} file to your SD card boot partition.

We also need a few features to be enabled in the U-Boot SPL. So start
U-Boot's \code{make menuconfig}:
\begin{itemize}
\item In the \code{Environment menu}, unselect \code{SPL environment is
      not stored} (\projconfigval{u-boot}{CONFIG_SPL_ENV_IS_NOWHERE}{n}), and
      select \code{SPL Environment is in a FAT filesystem}
      (\projconfigval{u-boot}{CONFIG_SPL_ENV_IS_IN_FAT}{y}). This way, we
      can have a \code{boot_os} environment variable that if set to
      \code{yes}, will allow to boot Linux directly. See the
      \projfunc{u-boot}{spl_start_uboot} function in the
      \projfile{u-boot}{board/ti/am335x/board.c} file for details. A
      faster option to skip reading the environment would be to
      modify this function so that it always returns \code{0}.
\item In the \code{SPL / TPL} menu, enable {\em Support SPL loading
      and booting of Legacy images}
      (\projconfigval{u-boot}{CONFIG_SPL_LEGACY_IMAGE_SUPPORT}{y}).
      That's needed to support loading the legacy \code{uImage} file.
\end{itemize}

Also make sure that \projconfigval{u-boot}{CONFIG_SPL_OS_BOOT}{y}
({\em Activate Falcon Mode}).
That's the case for our current configuration, but configurations
for other boards may not have this by default.

Compile U-Boot again
and copy the \code{u-boot.img} and \code{MLO} files to the boot
partition too.

Now, let's run the final preparation step. We will set the
\code{bootargs} environment variable, load the kernel and DTB, and
use U-Boot's \code{spl export} command to prepare a ready to boot record
with the DTB contents, the \code{bootargs}, the kernel loading addreses
and other information that Linux would need to boot. Note that the
U-Boot SPL will still load the \code{uImage} file from the FAT
filesystem in the first partition of the SD card.

In the below command, you'll see that we can use U-Boot's ready made
\code{loadaddr} and \code{fdtaddr} variables for addresses where to load
the kernel and DTB. At least this works with U-Boot for our board.

\begin{verbatim}
load mmc 0:1 ${loadaddr} uImage
load mmc 0:1 ${fdtaddr} dtb
setenv bootargs console=ttyS0,115200n8 rdinit=/playvideo
spl export fdt ${loadaddr} - ${fdtaddr}
\end{verbatim}

You can then see that \code{spl export} prepared everything to boot the
Linux kernel with the provided DTB, but didn't do it. At the end, it
tells you where the exported data were stored in RAM:

\begin{verbatim}
## Booting kernel from Legacy Image at 82000000 ...
   Image Name:   Linux-5.11.11-dirty
   Created:      2021-04-13   9:48:35 UTC
   Image Type:   ARM Linux Kernel Image (uncompressed)
   Data Size:    2842016 Bytes = 2.7 MiB
   Load Address: 80008000
   Entry Point:  80008000
   Verifying Checksum ... OK
## Flattened Device Tree blob at 88000000
   Booting using the fdt blob at 0x88000000
   Loading Kernel Image
   Loading Device Tree to 8ffec000, end 8ffff3ff ... OK
subcommand not supported
subcommand not supported
   Loading Device Tree to 8ffd5000, end 8ffeb3ff ... OK
Argument image is now in RAM: 0x8ffd5000
WARN: FDT size > CMD_SPL_WRITE_SIZE
\end{verbatim}

The last thing to do is to store such information in an \code{args} file
in the FAT partition on the MMC. Through the \code{fdtargsaddr} and
\code{fdtargsaddr} environment variables, we know where to copy the
data from and their size:

\begin{verbatim}
fatwrite mmc 0:1 ${fdtargsaddr} args ${fdtargslen}
\end{verbatim}

You're ready to go and reboot your board with the SD card inside.
You should not longer see the U-Boot second stage being loaded, but just
the SPL and the kernel.

If this doesn't work yet, please ask your instructor for advice and help.

When it works, update your table again:

\begin{tabular}{| l | l | r |}
  \hline
  Step & Duration & Description \\
  \hline
  \hline
  U-Boot SPL & & Between \code{U-Boot SPL 2021.10} and \code{Starting kernel} \\
  \hline
  Kernel + Init scripts & & Between \code{Starting kernel} and \code{Starting ffmpeg} \\
  \hline
  Application & & Between \code{Starting ffmpeg} and \code{First frame decoded} \\
  \hline
  \hline
  Total & & \\
  \hline
\end{tabular}

\section{Going further}

There are several things we can do to try to further optimize things:

\begin{itemize}
\item Reduce the features of the U-Boot SPL to make it load faster,
      modify \projfunc{u-boot}{spl_start_uboot} so that it doesn't have
      to load an environment. Also try loading the \code{uImage} file
      directly from raw MMC, so that there is no FAT intermediate layer.
\item As our storage is now faster, it can be interesting to explore the
      various kernel compression schemes again. The optimum solution may be a
      different one.
\item Look for a solution to eliminate the delay detecting the USB
      webcam.
\item If you don't manage to get rid of this delay, at least take
      advantage of this spare time to show signs of life on the screen, by
      implementing a splashscreen. You can even implement an animation.
      One thing you can do is use BusyBox's \code{fbsplash} tool, to first
      show an image on the framebuffer, and then even show a progress bar
      (knowing how much time you have to wait for the camera to be ready).
\end{itemize}

