\subchapter{First Yocto build}{Your first dive into a Yocto project and its
	build mechanism}

During this lab, you will:
\begin{itemize}
  \item Setup a Yocto environment
  \item Configure the project and choose a target
  \item Build your first Yocto image
\end{itemize}

\section{Setup}

Before starting this lab, make sure your home directory is not encrypted. Yocto
cannot be used on top of an eCryptFS file system due to limitations on file name
lengths.

Go to the \code{\$HOME/felabs-yocto/} directory.

Install the required packages:
\begin{verbatim}
sudo apt-get install chrpath gawk texinfo
\end{verbatim}

\input{../labs/common/setup-ethernet-communication.tex}

\section{Download Yocto}

Download the latest stable version of the Yocto project and extract it:
\begin{verbatim}
wget http://downloads.yoctoproject.org/releases/yocto/yocto-1.5.1/\
  poky-dora-10.0.1.tar.bz2
tar xf poky-dora-10.0.1.tar.bz2
\end{verbatim}

Go to the Yocto root directory: \code{cd poky-dora-10.0.1}

Then download the Yocto TI layer:
\begin{verbatim}
git clone git://git.yoctoproject.org/meta-ti.git
\end{verbatim}

\section{Setup the build environment}

Export all needed variables and setup the build directory:
\begin{verbatim}
source oe-init-build-env
\end{verbatim}

In order to choose the target and to configure the generic build settings,
edit the local configuration file (\code{\$BUILDDIR/conf/local.conf}). Set
the target machine to \code{beaglebone} and update the parallelization variables
(\code{BB_NUMBER_THREADS} and \code{PARALLEL_MAKE}) according to your computer
capabilities.

Also, if you need to save disk space on your computer you can add \code{INHERIT
+= "rm_work"} in the previous configuration file. This will remove the
package work directory once a package is built.

Finally, don't forget to let the configuration aware of the TI layer. Edit the
layer configuration file (\code{\$BUILDDIR/conf/bblayers.conf}) and append the
full path to the \code{meta-ti} directory to the \code{BBLAYERS} variable.

\section{Build your first image}

Now that you're ready to start the compilation, simply run:
\begin{verbatim}
bitbake core-image-minimal
\end{verbatim}

Once the build finished, all output images can be find under
\code{\$BUILDDIR/tmp/deploy/images/beaglebone}.

\section{Setup the internal eMMC}

In order to boot from the internal eMMC, we should make sure it is properly
setted up before writing our images.

\section{Flash the board}

In order to boot the BeagleBone Black we will flash the bootloader and the
kernel image in the internal eMMC here.

On the training computer copy the bootloader, kernel and rootfs images in the
root TFTP directory:
\begin{verbatim}
cp $BUILDDIR/tmp/deploy/images/beaglebone/{MLO,u-boot.img,zImage} /tftp
\end{verbatim}

Then retrieve these files on the BeagleBone Board memory. We will first download
the first stage bootloader, the \code{MLO}:
\begin{verbatim}
tftp 0x80200000 MLO
\end{verbatim}

Finally flash this file into the internal eMMC:
\begin{verbatim}
TBD
\end{verbatim}
