\subchapter{Build the system}{Objective: compile the root filesystem.}

After this lab, you will have a ready to use root filesystem to boot
your system with, including a video player application.

We haven't compiled the bootloader and kernel for our board yet,
but since this part can take a long time (especially compiling the
cross-compiling toolchain), let's start it now, while we are still
fetching kernel sources or going through lectures.

\section{Setup}

As specified in the Buildroot
manual\footnote{\url{https://buildroot.org/downloads/manual/manual.html\#requirement-mandatory}},
Buildroot requires a few packages to be installed on your
machine. Let's install them using Ubuntu's package manager:

\begin{verbatim}
sudo apt install sed make binutils gcc g++ bash patch \
  gzip bzip2 perl tar cpio python3 unzip rsync wget libncurses-dev
\end{verbatim}

You will later also find that you also need extra packages:

\begin{verbatim}
sudo apt install bison flex
\end{verbatim}

\section{Choosing a Buildroot release}

Buildroot is one of the best tools for building a custom root filesystem
for a dedicated embedded system with a fixed set of features, typically
like the one we're trying to build.

Go to the \code{~/__SESSION_NAME__-labs/rootfs/buildroot/} directory.

We will use the latest revision of the \code{2022.02} release, which is
one of Buildroot's long term releases:

\begin{verbatim}
git tag | grep 2022.02
git checkout 2022.02.12
\end{verbatim}

\section{Configuring Buildroot}

To minimize external dependencies and maximize flexibility, we will ask
Buildroot to generate its own toolchain. This can be better than using
external toolchains, as we have the ability to tweak toolchain settings
in a fine way if needed.

Start the Buildroot configuration utility:

\begin{verbatim}
make menuconfig
\end{verbatim}

\begin{itemize}

\item \code{Target Options} menu
  \begin{itemize}

  \item \code{Target Architecture}: select \code{ARM (little endian)}

  \item \code{Target Architecture Variant}: select \code{cortex-A8}

  \item On ARM two {\em Application Binary Interfaces} are available:
    \code{EABI} and \code{EABIhf}. Unless you have backward
    compatibility concerns with pre-built binaries, \code{EABIhf} is
    more efficient, so make this choice as the \code{Target ABI}
    (which should already be the default anyway).

  \item The other parameters can be left to their default value:
    \code{ELF} is the only available \code{Target Binary Format},
    \code{VFPv3-D16} is a sane default for the {\em Floating Point
      Unit}, and using the \code{ARM} instruction set is also a good
    default (we will later try the \code{Thumb-2} instruction set for
    slightly more compact code).
  \end{itemize}

\item \code{Build options} menu
  \begin{itemize}
  \item Enable \code{Enable compiler cache}
  \end{itemize}

\item \code{Toolchain} menu
  \begin{itemize}
  \item \code{Kernel Headers}: select \code{Manually specified Linux version}
	and then set \code{linux version} to \code{6.1}. This corresponds
	to the Linux kernel version we are going to use. For the system
	we want to build, which doesn't need system calls introduced
	by recent kernels, we could have used an older version of the
	kernel headers.
  \item \code{GCC compiler Version}: select \code{gcc 11.x}. This allows
        to have the most recent version of the compiler supported by
	this Buildroot release, and the best available optimizations.
  \item Keep all the other settings unmodified. We will get
	a toolchain with the {\em uClibc} library this way.
  \end{itemize}

\item \code{System configuration} menu
  \begin{itemize}
  \item Unselect \code{remount root filesystem read-write during boot}.
    This way, we will keep the root filesystem in read-only mode. When
    we make tests and reboot the system multiple times, this avoids
    filesystem recovery with approximately takes 4 seconds and adds
    jitter to our measurements.
  \end{itemize}

\item \code{Target packages} menu
  \begin{itemize}
  \item In \code{Audio and video applications}, select \code{ffmpeg} and
  inside the \code{ffmpeg} submenu, add the below options:
     \begin{itemize}
     \item Select \code{Build libswscale}
     \end{itemize}
  \end{itemize}

\item For the moment, all the remaining default settings are fine for us.
\end{itemize}

However, we need to do one thing to customize the root filesystem: we
need to add a script that will automatically start the \code{ffmpeg}
video player.

To do so, we will use Buildroot's {\em Root filesystem overlay}
capability, which allows to drop ready-made files into the final root
filesystem archive\footnote{See
\url{https://buildroot.org/downloads/manual/manual.html\#customize} for
details.}.

To begin with, let's start by creating a specific directory to store our
Buildroot customizations for our project.

\begin{verbatim}
mkdir board/beaglecam
\end{verbatim}

And in this directory, let's create a directory for root filesystem
overlays:

\begin{verbatim}
mkdir board/beaglecam/rootfs-overlay
\end{verbatim}

Now, let's copy a script that we're providing you to \code{etc/init.d}:

\begin{verbatim}
mkdir -p board/beaglecam/rootfs-overlay/etc/init.d/
cp ~/__SESSION_NAME__-labs/rootfs/data/S50playvideo board/beaglecam/rootfs-overlay/etc/init.d/
\end{verbatim}

We can now run \code{make menuconfig} again and in \code{System
configuration}, add \code{board/beaglecam/rootfs-overlay} to
\code{Root filesystem overlay directories}.

\section{Running Buildroot}

We are now ready to execute Buildroot:

\begin{verbatim}
make
\end{verbatim}

Enjoy, and be patient, as building a cross-compiling toolchain takes
time!
