\subchapter
{Kernel debugging: OOPS analysis and KGDB}
{Objectives:
  \begin{itemize}
	  \item Analyzing an {\em oops}.
	  \item Debugging with {\em KGDB}.
  \end{itemize}
}

\section{OOPS analysis}
We noticed that the watchdog command generated a crash on the kernel. In order
to reproduce the crash, run the following command on the target:

\begin{bashinput}
$ watchdog -T 10 -t 5 /dev/watchdog0
\end{bashinput}

Immediatly after executing this commands, you'll see that the kernel will report
an OOPS!

\subsection{Analyzing the crash message}

Analyze the crash message carefully. Knowing that on ARM, the \code{PC}
register contains the location of the instruction being executed, find
in which function does the crash happen, and what the function call
stack is.

Using Elixir (\url{https://elixir.bootlin.com/linux/latest/source}) or the
kernel source code, have a look at the definition of this function. In most
cases, a careful review of the driver source code is enough to understand the
issue. But not in that case!

\subsection{Locating the exact line where the error happens}

Even if you already found out which instruction caused the crash, it's
useful to use information in the crash report.

If you look again, the report tells you at what offset in the function
this happens. We will disassemble the code for this function to understand
exactly where the issue happened.

That is where we need a kernel compiled with \kconfig{CONFIG_DEBUG_INFO},
which is already enabled in the kernel we compiled in the initial lab.
This way, the kernel vmlinux file is compiled with \code{-g} compiler flag,
which adds a lot of debugging information (matching between source code
lines and assembly for instance).

Using \code{addr2line}, find the exact source code line were the crash happened.
For that, you can use the following command on your development host:

\begin{bashinput}
$ addr2line -e /home/$USER/__SESSION_NAME__-labs/buildroot/output/build/linux-%\workingkernel%/vmlinux -a <crash_address>
\end{bashinput}

This can also be done automatically using \code{decode_stacktrace.sh}. First,
copy/paste the OOPS message into the \code{~/__SESSION_NAME__-labs/oops.txt} file.
Then, using the script provided by the kernel, execute the following command:

\begin{bashinput}
$ cd /home/$USER/__SESSION_NAME__-labs/buildroot/output/build/linux-%\workingkernel%/
$ export ARCH=%\arch%
$ export CROSS_COMPILE=/home/$USER/__SESSION_NAME__-labs/buildroot/output/host/bin/%\crosscompile%
$ ./scripts/decode_stacktrace.sh vmlinux < ~/__SESSION_NAME__-labs/oops.txt
\end{bashinput}

We can even go a step further and use the cross GDB to open vmlinux and
locate the function and corresponding offset in assembly

\begin{bashinput}
$ ${CROSS_COMPILE}gdb /home/$USER/__SESSION_NAME__-labs/buildroot/output/build/linux-%\workingkernel%/vmlinux
(gdb) disassemble <function>
\end{bashinput}

\section{KGDB debugging}
In order to debug this OOPS, we'll use KGDB which is an in-kernel debugger.
The provided image already contains the necessary KGDB support and the watchdog
has been disabled to avoid rebooting while debugging. In order to use KGDB and
the console simultaneously, compile and run kdmx on your development host:

\begin{bashinput}
$ cd /home/$USER/__SESSION_NAME__-labs
$ git clone https://git.kernel.org/pub/scm/utils/kernel/kgdb/agent-proxy.git
$ cd agent-proxy/kdmx
$ make
$ ./kdmx -n -d -p/dev/ttyACM0 -b115200
serial port: /dev/ttyACM0
Initalizing the serial port to 115200 8n1
/dev/pts/7 is slave pty for terminal emulator
/dev/pts/8 is slave pty for gdb

Use <ctrl>C to terminate program
\end{bashinput}

Note: the slave ports number will depend on the run.

\textbf{Important: before using \code{/dev/pts/7} and \code{/dev/pts/8}, the
picocom process that did open \code{/dev/ttyACM0} must be closed!}

On the target, setup KGDB by setting the console to be used for that purpose in
kgdboc module parameters:

\begin{bashinput}
$ echo %\consoletty% > /sys/module/kgdboc/parameters/kgdboc
\end{bashinput}

Once done, trigger the crash by running the watchdog command, the system will
automatically wait for a debugger to be attached. Run the cross GDB and
attach a gdb process to KGDB with the following command:

\begin{bashinput}
$ ${CROSS_COMPILE}gdb ${HOME}/__SESSION_NAME__-labs/buildroot/output/build/linux-%\workingkernel%/vmlinux
(gdb) target remote /dev/pts/8
\end{bashinput}

{\em TIP: in order to allow auto-loading of python scripts, you can add
\code{set auto-load safe-path /} in your .gdbinit file}

First of all, confirm with GDB the information that were previously obtained
post-crash. This will allow you to also display variables values. Starting from that
point, we will add a breakpoint on the \code{watchdog_set_drvdata()} function.
However, this function is called early in boot so we will need to actually
attach with KGDB at boot time. To do so, we'll modify the bootargs to specify
that. In U-Boot, add the following arguments to bootargs using \code{env edit}:

\begin{bashinput}
=> env edit bootargs
=> <existing bootargs> kgdboc=%\consoletty%,115200 kgdbwait
=> boot
\end{bashinput}

Then the kernel will halt during boot waiting for a GDB process to be attached.
Attach gdb client from your development host using the same command that was previously
used:

\begin{bashinput}
$ ${CROSS_COMPILE}gdb /home/$USER/__SESSION_NAME__-labs/buildroot/output/build/linux-%\workingkernel%/vmlinux
(gdb) target remote /dev/pts/8
\end{bashinput}

Note: if you prefer using gdb-multiarch instead of the cross-gdb we have
built and do not specify a file to be used, gdb-multiarch won't be able to
detect the architecture automatically and the target command will fail. In
that case, you can set the architecture using:

\begin{bashinput}
(gdb) set arch %\arch%
(gdb) set gnutarget %\ifdefstring{\board}{beagleplay}{elf64}{elf32}%-littlearm
\end{bashinput}

Before continuing the execution, add a breakpoint on
\code{watchdog_set_drvdata()} using the \code{break} GDB command and then
continue the execution using the \code{continue} command

\begin{bashinput}
(gdb) break watchdog_set_drvdata
(gdb) continue
\end{bashinput}

Analyze the subsequent calls and find the place where the driver data are
clobbered.

TIP: you can fix the problem in "live" by modifying the content of the
\code{wdd->driver_data} variable directly using the following command:

\begin{bashinput}
(gdb) p/x var=hex_value
\end{bashinput}

Use it to set the variable with the previous value that was used before getting
clobbered with NULL. Once done, continue the execution and verify that you fixed
the problem using the \code{watchdog} command.

\ifdefstring{\board}{stm32mp1}{
{\em Note: In theory, we could have added a watchpoint to watch the address that
was modified but the ARM32 platform does not provide watchpoint support with
KGDB.}
}

\subsection{Debugging a module}

KGDB also allows to debug modules and thanks to the GDB python scripts
(\code{lx-symbols}) mainly, it is as easy as debugging kernel core code. In
order to test that feature, we are going to compile a test module and break on
it. On your development host, build the module:

\begin{bashinput}
$ cd /home/$USER/__SESSION_NAME__-labs/nfsroot/root/kgdb
$ export CROSS_COMPILE=/home/$USER/__SESSION_NAME__-labs/buildroot/output/host/bin/%\crosscompile%
$ export ARCH=arm
$ export KDIR=/home/$USER/__SESSION_NAME__-labs/buildroot/output/build/linux-%\workingkernel%/
$ make
\end{bashinput}

Then on the target, insert the module using insmod:
\begin{bashinput}
# cd /root/kgdb
# insmod kgdb_test.ko
\end{bashinput}

We can now enter KGDB mode and attach the external gdb to it. We will do
that using the magic SySrq 'g' key. Before that, ensure the tty
device for gdb already set, or set it now:

\begin{bashinput}
# echo %\consoletty% > /sys/module/kgdboc/parameters/kgdboc
# echo g > /proc/sysrq-trigger
\end{bashinput}

The kernel will then enter KGDB mode and will wait for a gdb connection. On the
development platform, start it and attach to the target:

\begin{bashinput}$
$ ${CROSS_COMPILE}gdb /home/$USER/__SESSION_NAME__-labs/buildroot/output/build/linux-%\workingkernel%/vmlinux
(gdb) target remote /dev/pts/8
\end{bashinput}

Then you will need to execute the \code{lx-symbols} command in gdb to reload the
symbols from the module. You'll also need to pass a list of path that contains
the external modules:

\begin{bashinput}
(gdb) lx-symbols /home/${USER}/__SESSION_NAME__-labs/nfsroot/root/kgdb/
loading vmlinux
scanning for modules in /home/<user>/__SESSION_NAME__-labs/nfsroot/root
loading @0xbf000000: /home/<user>/__SESSION_NAME__-labs/nfsroot/root/kgdb/kgdb_test.ko
\end{bashinput}

{\em NOTE: If KGDB was already connected and the lx scripts loaded, then
\code{lx-symbols} would be run automatically on module loading.}

Finally, add a breakpoint right after the \code{pr_debug()} call and continue
the execution to trigger it:

\begin{bashinput}
(gdb) break kgdb_test.c:17
(gdb) continue
\end{bashinput}

At some point, the breakpoint will be triggered. Try to display the variable
\code{i} to display the current loop value.

Note: Due to a GDB bug, sometimes, gdb will crash when continuing. You can
use a temporary breakpoint using the gdb \code{tbreak} command to workaround
this problem.

Bonus: as a side quest you can try to enable the \code{pr_debug()} call using
the dynamic debug feature of the kernel. This can be done using the
\code{/proc/dynamic_debug/control} file.
