\section{Cloning the mainline Linux tree}

To begin working with the Linux kernel sources, we need to clone its
reference git tree, the one managed by Linus Torvalds.

However, this requires downloading more than 2.8 GB of data. If you
are running this command from home, or if you have very fast access to
the Internet at work (and if you are not 256 participants in the
training room), you can do it directly by connecting to
\url{https://git.kernel.org}:

\begin{verbatim}
git clone https://git.kernel.org/pub/scm/linux/kernel/git/torvalds/linux
cd linux
\end{verbatim}

If Internet access is not fast enough and if multiple people have to
share it, your instructor will give you a USB flash drive with a
\code{tar.gz} archive of a recently cloned Linux source tree.

You will just have to extract this archive in the current directory,
and then pull the most recent changes over the network:

\begin{verbatim}
tar xf linux-git.tar.gz
cd linux
git checkout master
git pull
\end{verbatim}

Of course, if you directly ran
\code{git clone}, you won't have to run \code{git pull}, as
\code{git clone} already retrieved the latest changes. You may need to
run \code{git pull} in the future though, if you want to update a
newer Linux version.

\section{Accessing stable releases}

The Linux kernel repository from Linus Torvalds contains all the main
releases of Linux, but not the stable versions: they are maintained by
a separate team, and hosted in a separate repository.

We will add this separate repository as another {\em remote} to be
able to use the stable releases:

{\small
\begin{verbatim}
git remote add stable https://git.kernel.org/pub/scm/linux/kernel/git/stable/linux
git fetch stable
\end{verbatim}
}

As this still represents many git objects to download (2.4 GiB when 6.9 was
the latest version), if you are using an already downloaded git tree,
your instructor will probably have fetched the {\em stable} branch ahead
of time for you too. You can check by running:

\begin{verbatim}
git branch -a
\end{verbatim}

We will choose a particular stable version in the next labs.

Now, let's continue the lectures. This will leave time for the commands
that you typed to complete their execution (if needed).
