\subchapter{Use ADB}{Learn to use the Android Debug Bridge}

After this lab, you will be able to use ADB to:
\begin{itemize}
  \item Debug your system and applications
  \item Get a shell on a device
  \item Exchange files with a device
  \item Install new applications
\end{itemize}

\section{Setup}

Stay in the the \code{/home/<user>/felabs/android/linaro} directory.

\section{Get ADB}

ADB is distributed by Google as part of their SDK for Android. First,
you need to go to
\url{http://developer.android.com/sdk/}. Get the SDK for
Linux, and extract the archive that you downloaded.

Now, run the \code{tools/android} script, and in the \code{Packages}
window, select \code{Android SDK Platform-tools} under \code{Tools},
unselect all the other packages, and click on the \code{Install 1 package...} button.
Once you are done, you have \code{adb} installed in \code{./platform-tools/adb}.

However, the Android source code also has an embedded SDK, so you
already have the \code{adb} binary in the \code{out/host/linux-x86/bin/}.

To add this directory to your \code{PATH} after the Linaro build 
during the previous lab, you can go through the same environment setup steps
as earlier:

\begin{verbatim}
source build/envsetup.sh
lunch
\end{verbatim}

... and choose \code{beagleboard-eng} in the \code{lunch} menu.

\section{Disable USB suspend}

The kernel we have built uses USB suspend to save power. This will
be painful during the next labs, starting with this one.

To do so, go to the \code{kernel} directory. First, use the
\code{android_omap3_defconfig} file as an initial configuration. Load it using
\code{ARCH=arm make android_omap3_defconfig} and then modify it using
\code{ARCH=arm make menuconfig} to disable \code{CONFIG_USB_SUSPEND}
and enable \code{CONFIG_USB_OTG_WAKELOCK}. We will also need to enable
OHCI\footnote{USB Open Host Controller Interface, supporting only USB 1.1}
support to use the missile launcher.

Once you're done, exit and the configuration tool will save the new
file as \code{.config}. You can save it to a defconfig file by
running \code{ARCH=arm make savedefconfig}. Then, copy the newly created
\code{defconfig} file to the \code{arch/arm/configs} directory with
the \code{android_devkit8000_defconfig} name.

Once this is done, change the BeagleBoard product definition so that
it uses this new file.

\section{Configure USB access on your machine}

If you execute any command right now, you are very likely to have a
permission denied error. This is because the USB device associated to
the DevKit8000 doesn't have the right permissions to let you open it.

We need to make sure that \code{udev} sets the right permissions so that we can access
it as a user. To do so, create the file
\code{/etc/udev/rules.d/51-android.rules} and copy the following line:

\code{SUBSYSTEM=="usb", ATTR{idVendor}=="18d1", ATTR{idProduct}=="0001", MODE="0600", OWNER="<username>"}

Now, plug and unplug the DevKit8000, and you should now be able to use
\code{adb} from your user.

\section{Get logs from the device}

ADB provides many useful features to work with an existing Android device.
The first we will see is how to get the system logs from the whole system. To do
this, just run \code{adb logcat}.

You will see the device logs on your terminal. This is a huge amount of
information though, and it is difficult to find your way in all these lines.

The first thing we can do is download a little wrapper to \code{adb} to provide colored
logs. You can find it here: \url{http://jsharkey.org/downloads/coloredlogcat.pytxt}. Once downloaded, just run it and you will see the logs colored and
formatted to be easily readable.

ADB also provides filters to have a clearer output. These are formatted by the
\code{Tag:Priority} syntax. For example, if you want all logs from \code{MyApp},
just run \code{adb logcat MyApp:*}.

Now try to save all logs from Dalvik to a file using only \code{adb}.

\section{Get a shell on the device}

Having a shell on the device can prove pretty useful. ADB provides such a feature,
even though the Android-embedded shell is quite minimal. To access this shell,
just type \code{adb shell}.

You can also run a command directly on the device thanks to \code{adb shell}. To do
this, just append the command to run at the end. Now, try to get all the mounted
filesystems on the device.

\section{Push/Pull files to a device}

In the same way, ADB allows you to retrieve files from and send them to the
connected device, using the \code{push} and \code{pull} commands. 

