\subchapter{Secure boot: kernel FIT verification}{
	Objective:  Learn how to setup the second stage of secure boot: verification of
the kernel signature by U-Boot.}

\section{Activating signature verification in U-Boot}

\subsection{Configuring signature verification}

We now want U-Boot to verify the kernel FIT image before it loads it.
There is one configuration option in U-Boot that needs to be enabled for this.

That option is actually already enabled in our build, you can see it by looking at
the following patch:
\begin{bashinput}
$ cat board_setup/meta-security-training/recipes-bsp/u-boot/u-boot-kiss/freiheit93/0009-Add-FIT-signature.patch
commit 88fd6a65083c4e8cc2cfa2a38719a128fa8f3770
Author: Olivier Benjamin <olivier.benjamin@bootlin.com>
Date:   Sat May 9 14:45:09 2026 +0000

    Add FIT signature
    
    Signed-off-by: Olivier Benjamin <olivier.benjamin@bootlin.com>

diff --git a/configs/imx93_frdm_defconfig b/configs/imx93_frdm_defconfig
index 1deb7957cbc..8aef68c2f0b 100644
--- a/configs/imx93_frdm_defconfig
+++ b/configs/imx93_frdm_defconfig
@@ -132,6 +132,8 @@ CONFIG_BZIP2=y
 CONFIG_UTHREAD=y
 
 CONFIG_FIT=y
+CONFIG_FIT_SIGNATURE=y
+CONFIG_DEVICE_TREE_INCLUDES="secure-boot.dtsi"
 
 CONFIG_ENV_IS_NOWHERE=y
 CONFIG_ENV_IS_IN_MMC=y

\end{bashinput}

\begin{bashinput}
\end{bashinput}

\subsection{Passing the signing key to U-Boot}

Our build does not actually sign the FIT image, but U-Boot still boots it.
This shows that activating FIT signature is not enough for U-Boot to refuse to boot unsigned or wrongly signed 
FIT images.
It will only do so if it has an embedded key, and a list of images or configurations that \textbf{require} signature.
U-Boot expects this configuration in the form of a DTSI file, that can be included in its Device Tree Blob (DTB).
We can use the \code{key2dtsi.py} to generate the appropriate configuration based on a key.

As you can see, the patch above is already prepared to include this, in the form of \code{secure-boot.dtsi}.
Looking at that file, which is located in
\code{board_setup/meta-security-training/recipes-bsp/u-boot/u-boot-kiss/freiheit93/}.

You can now use \code{key2dtsi.py} to replace that file. You can use the version that Yocto deployed
as part of the native build of the \code{u-boot tools}:

\begin{bashinput}
(.venv) $ pip install pycryptodomex
(.venv) $ export UBOOT_TOOLS_DIR=build/tmp-glibc/work/x86_64-linux/u-boot-tools-native/2024.01
(.venv) $ ${UBOOT_TOOLS_DIR}/git/tools/key2dtsi.py --help
\end{bashinput}

\subsection{Testing}

Once this is done, you can rebuild U-Boot and flash it to the micro SD card.
Because Yocto does a lot of caching, the easiest way is probably to rebuild the entire image:

\begin{bashinput}
$ bitbake security-training-image
\end{bashinput}

Once it has been rebuilt, you can flash it to your micro SD card:
\begin{bashinput}
$ bmaptool copy build/tmp-glibc/deploy/images/freiheit93/security-training-image-freiheit93.rootfs.wic /dev/sdc
\end{bashinput}

Once you boot, this is the result you should see:

\begin{bashinput}
Verifying Hash Integrity ... Failed to verify required signature 'key-fit_signing_key'
error!
Unable to verify required signature for '' hash node in 'kernel-1' image node
Bad Data Hash
ERROR -2: can't get kernel image!
u-boot=>
\end{bashinput}

The system refuses to boot the FIT image, because it has not been signed. This is secure boot working as it
should. In the next step, we’ll build a signed FIT image that can be booted.

\section{Signing the FIT image}

We need to craft a new FIT image, properly signed for U-Boot to accept it.
We could do this directly in Yocto, but we'll do it manually here to learn the process.

FIT images, just like for AHAB, are a type of "container", so our first step will be identifying what components
need to go in.
In this case, our FIT image will contain:
\begin{itemize}
  \item our kernel: \code{build/tmp-glibc/work/freiheit93-oe-linux/linux-kiss/6.16/build/linux.bin}
  \item its DTB: \code{build/tmp-glibc/deploy/images/freiheit93/imx93-11x11-frdm.dtb}
\end{itemize}

In the labs-data, we included \code{unsigned_fit.its}, which a FIT specification similar to the one
that Yocto is currently using. Modify it to create your specification including the signature.
You can now use \code{mkimage} to create the signed FIT:

\begin{bashinput}
$ sudo apt install -y device-tree-compiler
(.venv) $ ${UBOOT_TOOLS_DIR}/build/tools/mkimage --help
\end{bashinput}

Once you're done, and you have flashed you new signed FIT image, your U-Boot boot log should look like this:

\begin{bashinput}
## Checking Image at 83000000 ...
   FIT image found
   FIT description: Kernel fitImage for security training Linux/6.16/freiheit93
    Image 0 (kernel-1)
     Description:  Linux kernel
     Type:         Kernel Image
     Compression:  uncompressed
     Data Start:   0x830000ec
     Data Size:    20439552 Bytes = 19.5 MiB
     Architecture: AArch64
     OS:           Linux
     Load Address: 0x80400000
     Entry Point:  0x83000000
     Hash algo:    sha256
     Hash value:   654f25e2b7aa16277b871ec3c39adc2d7da6bd69ce59fbb59a5726d431ade708
     Sign algo:    sha256,rsa4096:key-fit_signing_key
     Sign value:   c57a3f0461f3ca2d64d0a4f14...cd7f1527375a55e55117b227ceb40e8b131268
    Image 1 (fdt-imx93-11x11-frdm.dtb)
     Description:  Flattened Device Tree blob
     Type:         Flat Device Tree
     Compression:  uncompressed
     Data Start:   0x8437e698
     Data Size:    38662 Bytes = 37.8 KiB
     Architecture: AArch64
     Hash algo:    sha256
     Hash value:   c3e8f00b2b1c59b447a59a821e4b46c27270ab92bb14aa312236f2673b187b7a
     Sign algo:    sha256,rsa4096:key-fit_signing_key
     Sign value:   5732c710e63f5dac330bd5f9abf...ce31baf45895f9bc951c4e96e9ec78c317f
    Default Configuration: 'conf-imx93-11x11-frdm.dtb'
    Configuration 0 (conf-imx93-11x11-frdm.dtb)
     Description:  1 Linux kernel, FDT blob
     Kernel:       kernel-1
     FDT:          fdt-imx93-11x11-frdm.dtb
     Hash algo:    sha256
     Hash value:   unavailable
     Sign algo:    sha256,rsa4096:key-fit_signing_key
     Sign value:   2d2c1444d2094a9078ac7f35be446...bb136b3ca2a972a339afd0cbbc8e36cb14
## Checking hash(es) for FIT Image at 83000000 ...
   Hash(es) for Image 0 (kernel-1): sha256,rsa4096:key-fit_signing_key+ sha256+ 
   Hash(es) for Image 1 (fdt-imx93-11x11-frdm.dtb): sha256,rsa4096:key-fit_signing_key+ sha256+ 
## Loading kernel (any) from FIT Image at 83000000 ...
   Using 'conf-imx93-11x11-frdm.dtb' configuration
   Verifying Hash Integrity ... OK
   Trying 'kernel-1' kernel subimage
     Description:  Linux kernel
     Type:         Kernel Image
     Compression:  uncompressed
     Data Start:   0x830000ec
     Data Size:    20439552 Bytes = 19.5 MiB
     Architecture: AArch64
     OS:           Linux
     Load Address: 0x80400000
     Entry Point:  0x83000000
     Hash algo:    sha256
     Hash value:   654f25e2b7aa16277b871ec3c39adc2d7da6bd69ce59fbb59a5726d431ade708
     Sign algo:    sha256,rsa4096:key-fit_signing_key
     Sign value:   c57a3f0461f3ca2d64d0a4f14bd34...cd7f1527375a55e55117b227ceb40e8b131268
   Verifying Hash Integrity ... sha256,rsa4096:key-fit_signing_key+ sha256+ OK
\end{bashinput}
