\begin{frame}
  \frametitle{PM Building Blocks}
  \begin{itemize}
  \item Several power management \emph{building blocks}
    \begin{itemize}
    \item Suspend and resume
    \item CPUidle
    \item Runtime power management
    \item Frequency and voltage scaling
    \item Applications
    \end{itemize}
  \item Independent \emph{building blocks} that can be improved
    gradually during development
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Clock Framework (1)}
  \begin{itemize}
  \item Generic framework to manage clocks used by devices in the
    system
  \item Allows to reference count clock users and to shutdown the
    unused clocks to save power
  \item Simple API described in
    \url{http://free-electrons.com/kerneldoc/latest/DocBook/kernel-api/clk.html}
    \begin{itemize}
    \item \code{clk_get()} to get a reference to a clock
    \item \code{clk_enable()} to start the clock
    \item \code{clk_disable()} to stop the clock
    \item \code{clk_put()} to free the clock source
    \item \code{clk_get_rate()} to get the current rate
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Clock Framework (2)}
  \begin{itemize}
  \item The clock framework API and the \code{clk} structure are
    usually implemented by each architecture (code duplication!)
    \begin{itemize}
    \item See \code{arch/arm/mach-at91/clock.c} for an example
    \item This is also where all clocks are defined.
    \item Clocks are identified by a name string specific to a given
      platform
    \end{itemize}
  \item Drivers can then use the clock API. Example from
    \code{drivers/net/macb.c}:
    \begin{itemize}
    \item \code{clk_get()} called from the \code{probe()} function, to
      get the definition of a clock for the current board, get its
      frequency, and run \code{clk_enable()}.
    \item \code{clk_put()} called from the \code{remove()} function to
      release the reference to the clock, after calling
      \code{clk_disable()}
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
\frametitle{Clock Disable Implementation}
From \code{arch/arm/mach-at91/clock.c}: (2.6.36)
\begin{minted}[fontsize=\footnotesize]{c}
static void __clk_disable(struct clk *clk)
{
    BUG_ON(clk->users == 0);
    if (--clk->users == 0 && clk->mode)
        /* Call the hardware function switching off this clock */
        clk->mode(clk, 0);
    if (clk->parent)
        __clk_disable(clk->parent);
}

[...]

static void pmc_sys_mode(struct clk *clk, int is_on)
{
    if (is_on)
        at91_sys_write(AT91_PMC_SCER, clk->pmc_mask);
    else
        at91_sys_write(AT91_PMC_SCDR, clk->pmc_mask);
}
\end{minted}
\end{frame}

\begin{frame}
  \frametitle{Suspend and Resume}
  \begin{itemize}
  \item Infrastructure in the kernel to support suspend and resume
  \item Platform hooks
    \begin{itemize}
    \item \code{prepare()}, \code{enter()}, \code{finish()},
      \code{valid()} in a \code{platform_suspend_ops} structure
    \item Registered using the \code{suspend_set_ops()} function
    \item See \code{arch/arm/mach-at91/pm.c}
    \end{itemize}
  \item Device drivers
    \begin{itemize}
    \item \code{suspend()} and \code{resume()} hooks in the
      \code{*_driver} structures (\code{platform_driver},
      \code{usb_driver}, etc.)
    \item See \code{drivers/net/macb.c}
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Board-specific Power Management}
  \begin{itemize}
  \item Typically takes care of battery and charging management.
  \item Also defines \code{presuspend} and \code{postsuspend}
    handlers.
  \item Example: \code{arch/arm/mach-pxa/spitz_pm.c}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{arch/arm/mach-cpu/sleep.S}
  \begin{itemize}
  \item Assembly code implementing CPU specific suspend and resume
    code.
  \item Note: only found on arm, just 3 other occurrences in other
    architectures, with other paths.
  \item First scenario: only a suspend function. The code goes in
    sleep state (after enabling DRAM self-refresh), and continues with
    resume code.
  \item Second scenario: suspend and resume functions. Resume
    functions called by the bootloader.
  \item Examples to look at:
    \begin{itemize}
    \item \code{arch/arm/mach-omap2/sleep24xx.S} (1st case)
    \item \code{arch/arm/mach-pxa/sleep.S} (2nd case)
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Triggering Suspend}
  \begin{itemize}
  \item Whatever the power management implementation, CPU specific
    \code{suspend_ops} functions are called by the \code{enter_state}
    function.
  \item \code{enter_state} also takes care of executing the suspend
    and resume functions for your devices.
  \item The execution of this function can be triggered from
    userspace. To suspend to RAM:
    \begin{itemize}
    \item \code{echo mem > /sys/power/state}
    \end{itemize}
  \item Can also use the s2ram program from
    \url{http://suspend.sourceforge.net/}
  \item Read \code{kernel/power/suspend.c}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Runtime Power Management}
  \begin{itemize}
  \item According to the kernel configuration interface: \emph{Enable
      functionality allowing I/O devices to be put into energy-saving
      (low power) states at run time (or autosuspended) after a
      specified period of inactivity and woken up in response to a
      hardware-generated wake-up event or a driver's request.}
  \item New hooks must be added to the drivers:
    \code{runtime_suspend()}, \code{runtime_resume()},
    \code{runtime_idle()}
  \item API and details on \code{Documentation/power/runtime_pm.txt}
  \item See also Kevin Hilman's presentation at ELC Europe 2010:
    \url{http://elinux.org/images/c/cd/ELC-2010-khilman-Runtime-PM.odp}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Saving Power in the Idle Loop}
  \begin{itemize}
  \item The idle loop is what you run when there's nothing left to run
    in the system.
  \item Implemented in all architectures in
    \code{arch/<arch>/kernel/process.c}
  \item Example to read: look for \code{cpu_idle} in
    \code{arch/arm/kernel/process.c}
  \item Each ARM cpu defines its own \code{arch_idle} function.
  \item The CPU can run power saving HLT instructions, enter NAP mode,
    and even disable the timers (tickless systems).
  \item See also \url{http://en.wikipedia.org/wiki/Idle_loop}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Managing Idle}
  \begin{itemize}
  \item Adding support for multiple idle levels
    \begin{itemize}
    \item Modern CPUs have several sleep states offering different
      power savings with associated wake up latencies
    \item Since 2.6.21, the dynamic tick feature allows to remove the
      periodic tick to save power, and to know when the next event is
      scheduled, for smarter sleeps.
    \item CPUidle infrastructure to change sleep states
      \begin{itemize}
      \item Platform-specific driver defining sleep states and
        transition operations
      \item Platform-independent governors (ladder and menu)
      \item Available for x86/ACPI, not supported yet by all ARM cpus.
        (look for \code{cpuidle*} files under \code{arch/arm/})
      \item See \code{Documentation/cpuidle/} in kernel sources.
      \end{itemize}
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{PowerTOP}
  \begin{itemize}
  \item \url{http://www.lesswatts.org/projects/powertop/}
    \begin{itemize}
    \item With dynamic ticks, allows to fix parts of kernel code and
      applications that wake up the system too often.
    \item PowerTOP allows to track the worst offenders
    \item Now available on ARM cpus implementing CPUidle
    \item Also gives you useful hints for reducing power.
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Frequency and Voltage Scaling (1)}
  \begin{itemize}
  \item Frequency and voltage scaling possible through the
    \code{cpufreq} kernel infrastructure.
    \begin{itemize}
    \item Generic infrastructure: \code{drivers/cpufreq/cpufreq.c} and
      \code{include/linux/cpufreq.h}
    \item Generic governors, responsible for deciding frequency and
      voltage transitions
      \begin{itemize}
      \item \code{performance}: maximum frequency
      \item \code{powersave}: minimum frequency
      \item \code{ondemand}: measures CPU consumption to adjust frequency
      \item \code{conservative}: often better than
        \code{ondemand}. Only increases frequency gradually when the
        CPU gets loaded.
      \item \code{userspace}: leaves the decision to a userspace
        daemon.
      \end{itemize}
    \item This infrastructure can be controlled from
      \code{/sys/devices/system/cpu/cpu<n>/cpufreq/}
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Frequency and Voltage Scaling (2)}
  \begin{itemize}
  \item CPU support code in architecture dependent files.  Example to
    read: \code{arch/arm/plat-omap/cpu-omap.c}
  \item Must implement the operations of the \code{cpufreq_driver}
    structure and register them using \code{cpufreq_register_driver()}
    \begin{itemize}
    \item \code{init()} for initialization
    \item \code{exit()} for cleanup
    \item \code{verify()} to verify the user-chosen policy
    \item \code{setpolicy()} or \code{target()} to actually perform
      the frequency change
    \end{itemize}
  \item See \code{Documentation/cpu-freq/} for useful explanations
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{PM QoS}
  \begin{itemize}
  \item PM QoS is a framework developed by Intel introduced in 2.6.25
  \item It allows kernel code and applications to set their
    requirements in terms of
    \begin{itemize}
    \item CPU DMA latency
    \item Network latency
    \item Network throughput
    \end{itemize}
  \item According to these requirements, PM QoS allows kernel drivers
    to adjust their power management
  \item See \code{Documentation/power/pm_qos_interface.txt} and Mark
    Gross' presentation at ELC 2008
  \item Still in very early deployment (only 4 drivers in 2.6.36).
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Regulator Framework}
  \begin{itemize}
  \item Modern embedded hardware have hardware responsible for voltage
    and current regulation
  \item The regulator framework allows to take advantage of this
    hardware to save power when parts of the system are unused
    \begin{itemize}
    \item A consumer interface for device drivers (i.e users)
    \item Regulator driver interface for regulator drivers
    \item Machine interface for board configuration
    \item sysfs interface for userspace
    \end{itemize}
  \item Merged in Linux 2.6.27.
  \item See \code{Documentation/power/regulator/} in kernel sources.
  \item See Liam Girdwood's presentation at ELC 2008
    \url{http://free-electrons.com/blog/elc-2008-report\#girdwood}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{BSP Work for a New Board}
  \begin{itemize}
  \item In case you just need to create a BSP for your board, and your
    CPU already has full PM support, you should just need to:
    \begin{itemize}
    \item Create clock definitions and bind your devices to them.
    \item Implement PM handlers (suspend, resume) in the drivers for
      your board specific devices.
    \item Implement runtime PM handlers in your drivers.
    \item Implement board specific power management if needed (mainly
      battery management)
    \item Implement regulator framework hooks for your board if
      needed.
    \item All other parts of the PM infrastructure should be already
      there: suspend / resume, cpuidle, cpu frequency and voltage
      scaling.
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Useful Resources}
  \begin{itemize}
  \item \code{Documentation/power/} in the Linux kernel sources.
    \begin{itemize}
    \item Will give you many useful details.
    \end{itemize}
  \item \url{http://lesswatts.org}
    \begin{itemize}
    \item Intel effort trying to create a Linux power saving community.
    \item Mainly targets Intel processors.
    \item Lots of useful resources.
    \end{itemize}
  \item \url{http://wiki.linaro.org/WorkingGroups/PowerManagement/}
    \begin{itemize}
    \item Ongoing developments on the ARM platform.
    \end{itemize}
  \item Tips and ideas for prolonging battery life
    \begin{itemize}
    \item \url{http://j.mp/fVdxKh}
    \end{itemize}
  \end{itemize}
\end{frame}
