\section{Licensing}

\subsection{Managing licenses}

\begin{frame}[fragile]
  \frametitle{Tracking license changes}
  \begin{itemize}
    \item The license of an external project may change at some point.
    \item The \code{LIC_FILES_CHKSUM} tracks changes in the license
      files.
    \item If the license's checksum changes, the build will fail.
      \begin{itemize}
        \item The recipe needs to be updated.
      \end{itemize}
    \begin{block}{}
    \begin{minted}{c}
LIC_FILES_CHKSUM = "                                \
    file://COPYING;md5=...                          \
    file://src/file.c;beginline=3;endline=21;md5=..."
    \end{minted}
    \end{block}
    \item \code{LIC_FILES_CHKSUM} is mandatory in every recipe, unless
      \code{LICENSE} is set to \code{CLOSED}.
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Package exclusion}
  \begin{itemize}
    \item We may not want some packages due to their licenses.
    \item To exclude a specific license, use
      \code{INCOMPATIBLE_LICENSE}
      \item To exclude all GPLv3 packages:
      \begin{block}{}
      \begin{minted}{c}
INCOMPATIBLE_LICENSE = "GPLv3"
      \end{minted}
      \end{block}
    \item License names are the ones used in the \code{LICENSE}
      variable.
  \end{itemize}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Commercial licenses}
  \begin{itemize}
    \item By default the build system does not include commercial
      components.
    \item Packages with a commercial component define:
      \begin{block}{}
      \begin{minted}{c}
LICENSE_FLAGS = "commercial"
      \end{minted}
      \end{block}
    \item To build a package with a commercial component, the package
      must be in the \code{LICENSE_FLAGS_WHITELIST} variable.
    \item Example, \code{gst-plugins-ugly}:
      \begin{block}{}
      \begin{minted}[fontsize=\small]{c}
LICENSE_FLAGS_WHITELIST = "commercial_gst-plugins-ugly"
      \end{minted}
      \end{block}
  \end{itemize}
\end{frame}
