\ifdefempty{\trainer}{}{
  \input{../slides/first-slides/\trainer.tex}
}

\begin{frame}
\frametitle{Rights to copy}
   \scriptsize
   © Copyright 2004-\the\year, Bootlin \\
   {\bf License: Creative Commons Attribution - Share Alike 3.0} \\
   \url{https://creativecommons.org/licenses/by-sa/3.0/legalcode} \\

   You are free:
   \begin{itemize}
     \item to copy, distribute, display, and perform the work
     \item to make derivative works
     \item to make commercial use of the work
   \end{itemize}

   Under the following conditions:
   \begin{itemize}
     \item {\bf Attribution}. You must give the original author credit.
     \item {\bf Share Alike}. If you alter, transform, or build upon
           this work, you may distribute the resulting work only under
           a license identical to this one.
     \item For any reuse or distribution, you must make clear to others
           the license terms of this work.
     \item Any of these conditions can be waived if you get permission
           from the copyright holder.
    \end{itemize}

    Your fair use and other rights are in no way affected by the above.
    \vfill
    {\bf Document sources:}
    \url{https://github.com/bootlin/training-materials/} \\
\end{frame}

%% If the materials a generated for a real session, not for the website

\ifdefempty{\trainer}{}{
  \begin{frame}
  \frametitle{Electronic copies of these documents}
     \begin{itemize}
        \item Electronic copies of your particular version of the
              materials are available on:\\
              {\scriptsize \url{\sessionurl}} \\
        \item You can download and open these documents to follow
	      lectures and labs, to look for explanations given earlier
              by the trainer and to copy and paste text during labs.
        \item This specific URL will remaing available for a long time.
	      This way, you can always access the exact instructions
              corresponding to the labs performed in this session.
        \item If you are interested in the latest versions of our
	      training materials, visit the description of each
              course on \url{https://bootlin.com/training/}.
    \end{itemize}
  \end{frame}
}
