\begin{frame}
\frametitle{Free Electrons: not a training company}
  Free Electrons is an engineering company, not a training company
  \begin{itemize}
    \item Training is just one of our activities
    \item Whether they are directly employed by Free Electrons, or whether
          they are external developers that we know very well, all 
          our trainers are engineers first, with extensive on-the-job
          experience.
    \item Free Electrons engineers spend most of their time on technical projects,
          and share this experience through training sessions and by
          keeping our training materials up to date.
    \item All our trainers also spend a lot of time contributing to the user
          and developer community, by contributing to projects (such as the Linux kernel,
          Buildroot and Barebox), and/or by sharing technical information (through
          blog posts, training materials and talks at international
          conferences)
  \end{itemize}
\end{frame}

\begin{frame}
\frametitle{Free Electrons at a glance}
  \begin{itemize}
    \item Created in 2004
    \item Locations: Orange, Toulouse, Saint Etienne / Lyon (France)
    \item Serving customers all around the world \\
          See \url{http://free-electrons.com/company/customers/}
    \item Head count: 6 (September 2012) \\
	  Only Free Software enthousiasts!
    \item Focus: Embedded Linux, Linux kernel, Android
          Free Software / Open Source
          for embedded and real-time systems.
    \item Activities: development, training, consulting, technical
          support.
    \item Added value: get the best of the user and development
          community and the resources it offers.
  \end{itemize}
\end{frame}

\begin{frame}
\frametitle{Free Electrons: what's special}
  \begin{itemize}
    \item Engineers recruited in the heart of the embedded Linux
   	  developer community.
    \item We are very familiar with the best solutions the
          community offers to product developers.
    \item Contributing as much as possible to the community:
          code, documentation, knowledge sharing, financial support.
    \item Our engineers regularly go to the top technical
          conferences. We know other developers very well.
    \item Nothing proprietary in Free Electrons. Everything we
          produce for our company is shared and transparent
          (in particular training materials and even evaluations
          from all our training sessions!).
  \end{itemize}
\end{frame}

\begin{frame}
\frametitle{Our main services}
  \begin{itemize}
     \item Linux kernel and board support package development,
    	   to support new an custom hardware:
           bootloader, initialization, device drivers, power
           management...
     \item Linux kernel mainlining: integrate support for your
           hardware in the official Linux kernel sources 
     \item Android porting and customization
     \item System development and building environment.
           Buildroot, OpenEmbedded and Yocto support.
     \item System integration: choosing the best components
           and making a custom system. 
     \item Boot time reduction
  \end{itemize}
\end{frame}

\begin{frame}
\frametitle{Free Electrons on-line resources}
  \begin{itemize}
    \item All our training materials:\\
          \url{http://free-electrons.com/docs/}
    \item Technical blog:\\
          \url{http://free-electrons.com/blog/} 
    \item Quarterly newsletter:\\
 	  \url{http://lists.free-electrons.com/mailman/listinfo/newsletter}
    \item News and discussions (LinkedIn):\\
	  \url{http://linkedin.com/groups/Free-Electrons-4501089}
    \item Quick news (Twitter):\\
          \url{http://twitter.com/free_electrons}
  \end{itemize}
\end{frame}
